---
title: "IE 6600 Section 01 Group 25 Hackathon "
author: "Aryak Bodkhe, Rishika Reddy, Saurabh Borse"
date: "2023-02-08"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# FITNESS TRACKER

### INTRODUCTION and PROBLEM STATEMENT

The aim of this report is to analyze the data collected from a Fitness
tracker, which is a wearable device designed to track various aspects of
an individual's physical activity, such as steps taken, distance
traveled, calories burned, and heart rate. The data analyzed in this
report was collected over a specified period of time and provides
insights into the physical activity patterns of 33 individuals wearing
the device.

This report will present the findings of the data analysis, including
trends and patterns in the individuals' physical activities, as well as
any correlations between different aspects of the data. The report will
also highlight any notable achievements or changes in the individuals'
physical activities during the specified period of time.

This report is distributed in three sections.

-   **Section 1:** Analysis of users' daily activities

-   **Section 2:** Analysis of steps taken and sleep patterns

-   **Section 3:** Analysis of Calories burnt. Overall, this report aims
    to provide a comprehensive overview of users' physical activity as
    recorded by their Fitness tracker, and will serve as a valuable tool
    for understanding their fitness level, tracking their progress, and
    identifying areas for improvement.

```{r, echo=FALSE, warning=FALSE}
#libraries
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(janitor))
suppressPackageStartupMessages(library(lubridate))


```

```{r, echo=FALSE, warning=FALSE}
#Reading csv files
dailyAct <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv", na.strings = "")
dailyCal <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailyCalories_merged.csv", na.strings = "")
dailyInt <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailyIntensities_merged.csv", na.strings = "")
dailySteps <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailySteps_merged.csv", na.strings = "")
dailySleep <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv", na.strings = "")
weights <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv", na.strings = "")

```

```{r, echo=FALSE, warning=FALSE}
#Cleaning the data

#Cleaned the date columns to make them same and transformed them to the right data type, before merging the datasets.

dailyAct <- dailyAct %>% 
  rename(Date = ActivityDate) %>% 
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))

dailySleep <- dailySleep %>% 
  rename(Date = SleepDay) %>% 
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))

weights <- weights %>% 
  select(-LogId) %>% 
  mutate(Date = as.Date(Date, format = "%m/%d/%y")) %>% 
  mutate(IsManualReport = as.factor(IsManualReport))
```

```{r, echo=FALSE, warning=FALSE}
data <- merge(merge(dailyAct, dailySleep, by = c('Id', 'Date'), all = TRUE), weights, by = c('Id', 'Date'), all = TRUE)

merge_1 <- merge(dailyAct, dailyCal, by = c("Id","Calories"))
merge_2 <- merge(dailyInt, dailyInt, by = c("Id","ActivityDay","SedentaryMinutes", "LightlyActiveMinutes","FairlyActiveMinutes","VeryActiveMinutes", "SedentaryActiveDistance", "LightActiveDistance", "ModeratelyActiveDistance", "VeryActiveDistance"))

merge_daily <- merge(merge_1, merge_2, by = c("Id","ActivityDay","SedentaryMinutes", "LightlyActiveMinutes","FairlyActiveMinutes","VeryActiveMinutes", "SedentaryActiveDistance", "LightActiveDistance", "ModeratelyActiveDistance", "VeryActiveDistance"))
  
#select(-ActivityDay) %>% rename(Date = ActivityDate)

daily_data <- merge(merge_daily, dailySleep, by = "Id",all=TRUE)
```

```{r, echo=FALSE, warning=FALSE}
#Removing some extra variables in the final (merged) data

#Looking at the final dataset, we observed that the columns TotalDistances and TrackerDistances are identical. Based on this, one of columns is deleted for the analysis. We also deleted some other variables that aren't useful for our analysis.

data <- data %>% select(-c(TrackerDistance, LoggedActivitiesDistance, TotalSleepRecords, WeightPounds, Fat, BMI, IsManualReport))
```

```{r, echo=FALSE, warning=FALSE}
#Setting up a common theme for our plots
custom_theme <- function() {
  theme(
    panel.border = element_rect(colour = "black", 
                                fill = NA, 
                                linetype = 1),
    panel.background = element_rect(fill = "white", 
                                    color = 'grey50'),
    panel.grid.minor.y = element_blank(),
    axis.text = element_text(colour = "black", 
                             face = "italic", 
                             family = "Helvetica"),
    axis.title = element_text(colour = "black", 
                              family = "Helvetica"),
    axis.ticks = element_line(colour = "black"),
    plot.title = element_text(size=23, 
                              hjust = 0.5, 
                              family = "Helvetica"),
    plot.subtitle=element_text(size=16, 
                               hjust = 0.5),
    plot.caption = element_text(colour = "black", 
                             face = "italic", 
                             family = "Helvetica")
  )
}
```

### SECTION 1:

### Analysis of users' daily activity

**Q. What is the distribution of users' daily activity in a week?**

```{r, echo=FALSE, warning=FALSE}
data %>% 
  mutate(weekdays = weekdays(Date)) %>% 
  select(weekdays, TotalSteps) %>%
  mutate(weekdays = factor(weekdays, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))) %>% 
  drop_na() %>% 
  ggplot(aes(weekdays, TotalSteps, fill = weekdays)) +
  geom_boxplot() +
  custom_theme() +
  scale_fill_brewer(palette="Pastel") +
  theme(legend.position="none") +
  labs(
    title = "Users' activity by day",
    x = "Day of the week",
    y = "Steps",
  )
```

**Insight:** The above visualization plot shows all the 33 users' daily
activity by steps and we can observe that the difference between the
days of the week is minimal.

**Q: What is the percentage of users that are lightly active (steps in
minutes)?**

```{r, echo=FALSE, warning=FALSE}
pie_chart <- data %>% select(VeryActiveMinutes, 
         FairlyActiveMinutes, 
         LightlyActiveMinutes, 
         SedentaryMinutes) %>%
  summarise(across(everything(), list(sum))) %>% 
  gather(active_level, minutes) %>% 
  mutate(active_level = factor(active_level, 
                          labels = c('Moderate Activity','Light Activity',
                                     'Sedentary','Heavy Activity')))


fig <- plot_ly(pie_chart, labels = ~active_level, values = ~minutes, type = 'pie', textposition = 'inside', textinfo = 'label+percent',  marker = list(colors = c("lightsalmon","palegreen","lightblue4","darkgreen")))

fig <- fig %>% layout (title= 'Distribution of daily activities of users (steps)',
                      xaxis = list(showgrid = FALSE, zeroline = FALSE, shoqticklabels = FALSE),
                      yaxis = list(showgrid = FALSE, zeroline = FALSE, shoqticklabels = FALSE))
fig
```

**Insight:** This visualization chart shows the distribution of users'
daily activity(steps) in minutes. We observed that only 1.12% of users'
do moderate activity and only 1.74% of the users' do heavy activity
daily. Large amount of users (81.3%) are sedentary. 15.8% of the users
are lightly active.

**Q: What is the percentage of users that are lightly active (calories
burnt)?**

```{r, echo=FALSE,warning=FALSE}
#With this new table we can now visualize both the user type distribution and the calories burned for every user type:
daily_data <- data %>%
summarise(user_type = factor(case_when(
    SedentaryMinutes > mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Sedentary",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes > mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Lightly Active",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes > mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Fairly Active",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes > mean(VeryActiveMinutes) ~ "Very Active",
),levels=c("Sedentary", "Lightly Active", "Fairly Active", "Very Active")), Calories, .group=Id) %>%

drop_na()

pie_chart1 <- daily_data %>%
group_by(user_type) %>%
summarise(total = n()) %>%
mutate(totals = sum(total)) %>%
group_by(user_type) %>%
summarise(total_percent = total / totals)

fig1 <- plot_ly(pie_chart1, labels = ~user_type, values = ~total_percent, type = 'pie', textposition = 'inside', textinfo = 'label+percent',  marker = list(colors = c("lightsalmon","darkgreen","lightblue4","palegreen")))

fig1 <- fig1 %>% layout (title= 'Distribution of daily activities of users (calories burnt)',
                      xaxis = list(showgrid = FALSE, zeroline = FALSE, shoqticklabels = FALSE),
                      yaxis = list(showgrid = FALSE, zeroline = FALSE, shoqticklabels = FALSE))
fig1
```

**Insight:** Here, we are analyzing the percentage of calories burnt by
the users with different active levels. And it is observed that 57.2%
are sedentary based on the calories that they burnt. 32.4% of the users
are lightly active.

### SECTION 2 : 

### Analysis of the steps taken and sleep patterns of User

```{r,echo=FALSE, warning=FALSE}
#install.packages("ggrepel")
#Dataframes
df_daily_activity<- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv", na.strings = "")
df_sleep<- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv", na.strings = "")
df_hourly_step_count<- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/hourlySteps_merged.csv", na.strings = "")
```

```{r,echo=FALSE, warning=FALSE}
#head(df_daily_activity)
#head(df_sleep )
#head(df_hourly_step_count)
```

```{r,echo=FALSE, warning=FALSE}
#n_unique(df_daily_activity$Id)
#n_unique(df_sleep$Id)
#n_unique(df_hourly_step_count$Id)
```

```{r,echo=FALSE, warning=FALSE}
#sum(duplicated(df_daily_activity))
#sum(duplicated(df_sleep))
#sum(duplicated(df_hourly_step_count))
```

```{r,echo=FALSE, warning=FALSE}
df_daily_activity <- df_daily_activity %>%
  distinct() %>%
  drop_na()

df_sleep <- df_sleep %>%
  distinct() %>%
  drop_na()

df_hourly_step_count <- df_hourly_step_count %>%
  distinct() %>%
  drop_na()
```

```{r,echo=FALSE, warning=FALSE}
#clean_names(df_daily_activity)
df_daily_activity<- rename_with(df_daily_activity, tolower)

#clean_names(df_sleep)
df_sleep <- rename_with(df_sleep, tolower)

#clean_names(df_hourly_step_count)
df_hourly_step_count <- rename_with(df_hourly_step_count, tolower)
```

```{r,echo=FALSE, warning=FALSE}
df_daily_activity <- df_daily_activity %>%
  rename(date = activitydate) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y"))

df_sleep <- df_sleep %>%
  rename(date = sleepday) %>%
  mutate(date = as_date(date,format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))


```

```{r,echo=FALSE, warning=FALSE}
df_hourly_step_count<- df_hourly_step_count %>% 
  rename(date_time = activityhour) %>% 
  mutate(date_time = as.POSIXct(date_time,format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))


```

```{r,echo=FALSE, warning=FALSE}
#Merging the datasets
activity_sleep_merged <- merge(df_daily_activity, df_sleep, by=c ("id", "date"))

```

```{r,echo=FALSE, warning=FALSE}
timeoftheday_steps_sleep <- activity_sleep_merged %>%
  mutate(weekday = weekdays(date))

timeoftheday_steps_sleep$weekday <-ordered(timeoftheday_steps_sleep$weekday, levels=c("Monday", "Tuesday", "Wednesday", "Thursday",
"Friday", "Saturday", "Sunday"))

 timeoftheday_steps_sleep <-timeoftheday_steps_sleep%>%
  group_by(weekday) %>%
  summarize (daily_steps = mean(totalsteps), daily_sleep = mean(totalminutesasleep))


```

```{r,echo=FALSE, warning=FALSE}
df_hourly_step_count <- df_hourly_step_count %>%
  separate(date_time, into = c("date", "time"), sep= " ") %>%
  mutate(date = ymd(date)) 
  
#head(df_hourly_step_count)
```

**Q. What is the distribution of Steps of users throughout the day?**

```{r,echo=FALSE, warning=FALSE}
df_hourly_step_count %>%
  group_by(time) %>%
  summarize(average_steps = mean(steptotal)) %>%
  ggplot() +
  geom_col(mapping = aes(x=time, y = average_steps, fill = average_steps)) + 
  custom_theme() +
  scale_fill_gradient(low = "#66CC99", high = "tan1")+
  theme(axis.text.x = element_text(angle = 90))+
  labs(
    title = "Hourly steps throughout the day",
    x = "",y="",
    caption = 'Data Source: FitBit Fitness Tracker Data'
  )
  
```

**Insight :** The above visualization plot shows the average hourly
steps of all the 32 users in a particular day. From what we can observe
from this plot is that all the users in average are more active between
the time of 8am and 7pm. Majorly step count increases between the time
of 5pm and 8pm.

**Q. What is the sleep distribution of users throughout the day?**

```{r,echo=FALSE, warning=FALSE}
activity_sleep_merged %>% 
  select(totalminutesasleep) %>% 
  drop_na() %>% 
  mutate(sleep_quality = ifelse(totalminutesasleep <= 420, 'Less than 7h',
                         ifelse(totalminutesasleep <= 540, '7h to 9h', 
                         'More than 9h'))) %>%
  mutate(sleep_quality = factor(sleep_quality, 
                          levels = c('Less than 7h','7h to 9h',
                                     'More than 9h'))) %>% 
  ggplot(aes(x = totalminutesasleep, fill = sleep_quality)) +
  geom_histogram(position = 'dodge', bins = 30) +
  custom_theme() +
  scale_fill_brewer(palette='Pastel2')+
  #scale_fill_manual(values=c("tan1", "#66CC99", "lightcoral")) +
  theme(legend.position = c(.80, .80),
        legend.title = element_blank(),
        legend.spacing.y = unit(0, "mm"), 
        panel.border = element_rect(colour = "black", fill=NA),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black")) +
  labs(
    title = "Sleep distribution",
    x = "Time slept (minutes)",
    y = "Count",
    caption = 'Data Source: FitBit Fitness Tracker Data'
  )
```

**Insight :** Now here, we our analysing the sleep distribution of all
user in minutes. From this visualization we can determine how many users
have a sound sleep. Good and healthy sleep range must be between 7hr to
9hr range i.e, between 630 and 720 minutes.

However the above visualization shows that majority of users average
minutes of sleep, following a normal distribution. A majority of users
sleep approximately 320 to 530 minutes.

Getting deeper into the analysis, we come up with this question.

**Q. Is there a correlation between sleep time and distance travelled by
the fit bit users?**

```{r,echo=FALSE, warning=FALSE}
activity_sleep_merged %>% 
  select(id, totaldistance, totalminutesasleep) %>% 
  group_by(id) %>% 
  summarise_all(list(~mean(., na.rm=TRUE))) %>% 
  drop_na() %>% 
  mutate(id = factor(id)) %>% 
  ggplot() +
  custom_theme() +
  geom_bar(aes(x = id, y = totaldistance), ,stat = "identity", fill = '#66CC99', alpha = 0.7)+ 
  #scale_fill_manual(values=c("tan1", "#66CC99", "lightcoral")) +
  scale_fill_brewer(palette='Paired')+
  geom_point(aes(x = id, y = totalminutesasleep/60), color = 'tan1', size=2) +
  geom_segment(aes(x = id, xend = id, y = 0, yend = totalminutesasleep/60), color = 'tan1' ,group = 1) +
  scale_y_continuous(limits=c(0, 12), name = "Total Distance", 
  sec.axis = sec_axis(~.*60, name = "Sleep in minutes")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Average distance vs average sleep by user",
    x = "Users",
    caption = 'Data Source: FitBit Fitness Tracker Data'
  )
```

**Insight :** Now we want to determine whether the step counts has a
correlation with the sleep time and how basically it affects the sleep
time of the users. To determine this, correlation we have decided to
plot between the total distance travelled by the users in a day with
total sleep time of user in minutes.

What we can infer from this plot is that covering a greater distance
doesn't necessarily mean that the user is going to have a better sleep
(on average).

Let's confirm this theory with the next question.

**Q. Does the amount of steps taken affect the sleep quality of the
users?**

```{r,echo=FALSE,warning=FALSE}
activity_sleep_merged %>% 
  select(totalminutesasleep, totalsteps) %>% 
  mutate(sleep_quality = ifelse(totalminutesasleep <= 420, 'Less than 7h',
                         ifelse(totalminutesasleep <= 540, '7h to 9h', 
                         'More than 9h'))) %>% 
  mutate(active_level = ifelse(totalsteps >= 15000,'More than 15,000 steps',
                        ifelse(totalsteps >= 10000,'10,000 to 14,999 steps',
                        ifelse(totalsteps >= 5000, '5,000 to 9,999 steps',
                        'Less than 4,999 steps')))) %>% 
  select(-c(totalminutesasleep, totalsteps)) %>% 
  drop_na() %>% 
  group_by(sleep_quality, active_level) %>% 
  summarise(counts = n()) %>% 
  mutate(active_level = factor(active_level, 
                                levels = c('Less than 4,999 steps',
                                           '5,000 to 9,999 steps',
                                           '10,000 to 14,999 steps',
                                           'More than 15,000 steps'))) %>% 
  mutate(sleep_quality = factor(sleep_quality, 
                          levels = c('Less than 7h','7h to 9h',
                                     'More than 9h'))) %>% 
  ggplot(aes(x = sleep_quality, 
             y = counts, 
              fill = sleep_quality)) +
  geom_bar(stat = "identity") +
  custom_theme() +
  scale_fill_brewer(palette='Pastel2')+
  #scale_fill_manual(values=c("tan1", "#66CC99", "lightcoral")) +
  facet_wrap(~active_level, nrow = 1) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(strip.text = element_text(colour = 'black', size = 8)) +
  theme(strip.background = element_rect(fill = "beige", color = 'black'))+
  labs(
    title = "Sleep quality by steps",
    x = "Sleep quality",
    y = "Step Count",
    caption = 'Data Source: FitBit Fitness Tracker Data'
  )

```

**Insight:** Now as mentioned earlier we know that the good quality of
sleep have the time range between 7hr to 9hr.

As we could not infer much from the earlier graph. We here are trying to
focus on the best sleep quality and what we infer is that on average the
best sleep is achieve when the total steps during the day are between
5000 steps and 14,999 steps. The optimal steps range to have a good
sleep quality is between 5,000 and 9,999.

### SECTION 3: 

### Analysis of Calories burnt by users

Burning calories through exercise is one of the crucial part for the
healthy lifestyle. Fitness tracker are accurate enough to measure
calorie burns throughout the day. The number of calories burnt for each
user is definitive to understand users' health habits as well as
tracking the progress of their regimen.

```{r, echo=FALSE, warning=FALSE}
dailyActivity_merged <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
dailyCalories_merged <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailyCalories_merged.csv")
dailyIntensities_merged <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailyIntensities_merged.csv")
dailySteps_merged <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/dailySteps_merged.csv")
sleepDay_merged <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
weightLogInfo_merged <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
hourly_calories <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv")

#DATA operations:
  
#Here I join all the daily tables data and the sleep data. They share the Id and date columns so it easy to simply merge them by Id and then leave only one date column.
merge_1 <- merge(dailyActivity_merged, dailyCalories_merged, by = c("Id","Calories"))
merge_2 <- merge(dailyIntensities_merged, dailyIntensities_merged, by = c("Id","ActivityDay","SedentaryMinutes", "LightlyActiveMinutes","FairlyActiveMinutes","VeryActiveMinutes", "SedentaryActiveDistance", "LightActiveDistance", "ModeratelyActiveDistance", "VeryActiveDistance"))

merge_daily <- merge(merge_1, merge_2, by = c("Id","ActivityDay","SedentaryMinutes", "LightlyActiveMinutes","FairlyActiveMinutes","VeryActiveMinutes", "SedentaryActiveDistance", "LightActiveDistance", "ModeratelyActiveDistance", "VeryActiveDistance")) %>%
select(-ActivityDay) %>% rename(Date = ActivityDate)

daily_data <- merge(merge_daily, sleepDay_merged, by = "Id",all=TRUE) %>% drop_na() %>% select(-SleepDay, -TrackerDistance)

options(repr.plot.width=30)

#In order to better visualize the data I will group the user into four categories based on for which of their activity types they have more minutes, this will be very useful to quickly see patterns and visualize them:
data_by_usertype <- daily_data %>%
summarise(user_type = factor(case_when(
    SedentaryMinutes > mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Sedentary",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes > mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Lightly Active",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes > mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Fairly Active",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes > mean(VeryActiveMinutes) ~ "Very Active",
),levels=c("Sedentary", "Lightly Active", "Fairly Active", "Very Active")), Calories, .group=Id) %>%
drop_na()

```

**Q1. Does burning calories have correlation with the type of users
based on how active they are?**

```{r , echo=FALSE, warning=FALSE}

ggplot(data_by_usertype, aes(user_type, Calories, fill=user_type)) +
    geom_boxplot(alpha=0.5) +
    theme(legend.position="none") +
    custom_theme()+
    scale_fill_brewer(palette = "Pastel")+
    labs(title="Calories burned by the User type", x=NULL) +
    theme(legend.position="none", text = element_text(size = 20),plot.title = element_text(hjust = 0.5))

```

**Insight:** The users are categorized into sedentary, lightly, fairly
and very active based on the minutes of their activity throughout the
day. A box plot is a suitable graph to compare these user types based on
median, quartiles and outliers.The average calories burnt by the users
gradually increases starting from sedentary to very active individuals.
Surprisingly, light and fairly active individuals do not show much
difference in burning the calories however, the median calorie burnt for
fairly active individuals is less than the lightly active individuals.
Very active individuals are almost burning 3000 calories on an average
per day.

**Q2. Does burning calories have correlation with the number of
steps/distance that users travel?**

```{r, echo=FALSE, warning=FALSE}
daily_data %>%
summarise(
distance = factor(case_when(
    TotalDistance < 5 ~ "< 5 mi",
    TotalDistance >= 5 & TotalDistance <= 7 ~ "5 > & < 7 mi",
    TotalDistance > 7 ~ "> 7 mi",
),levels = c("> 7 mi","5 > & < 7 mi","< 5 mi")),
steps = factor(case_when(
    TotalSteps < 6000 ~ "< 6k steps",
    TotalSteps >= 6000 & TotalSteps <= 10000 ~ "6k > & < 10k Steps",
    TotalSteps > 10000 ~ "> 10k Steps",
),levels = c("> 10k Steps","6k > & < 10k Steps","< 6k steps")),
Calories) %>%
ggplot(aes(steps,Calories,fill=steps)) +
    geom_boxplot() +
    scale_fill_brewer(palette = "Pastel2")+
    custom_theme()+
    facet_wrap(~distance)+
    labs(title="Calories burned by Steps and Distance",x=NULL) +
    theme(legend.position="none", text = element_text(size = 20),plot.title = element_text(hjust = 0.5), 
          axis.text.x = element_text(size= 10, angle = 45, vjust = 1, hjust=1))+
  theme(strip.text = element_text(colour = 'black', size = 8)) +
  theme(strip.background = element_rect(fill = "beige", color = 'black'))

```

**Insight:** The users are categorized based on the distance traveled
(Less than 5 Miles, 5 to 7 miles and more than 7 miles) as well as
Number of steps taken (less than 6000, 6000 to 10000 and more than
10000). Based on the summary values, this Boxplot shows the calories
burned by the three categories of steps, which are then faceted by the
three categories of distance traveled. It determines which component is
more crucial to the number of calories burned, and it's interesting to
note that the most calories are burned by the "6,000 \> 10,000 Steps"
and "\> 7 miles," which would suggest some sort of running activity that
allows participants to cover more distance with fewer steps.

It is also intriguing to see that, the similarity between the "\> 10,000
Steps" in the mean distance section and the "6,000 Steps" in the low
distance section, which supports the notion that speed is the primary
determinant of calories burned.

**Q3. What hour of the day is the most popular in users to burn
calories?**

```{r,echo=FALSE}
hourly_calories <- read.csv("/Users/aryakbodkhe/Desktop/Comp Viz/Hackathon/Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv")
hourly_calories <- hourly_calories %>%
  distinct() %>%
  drop_na()

hourly_calories <- rename_with(hourly_calories, tolower)

hourly_calories <- hourly_calories %>%
  rename(date_time = activityhour) %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %I:%M:%S %p", tz = Sys.timezone()))%>%
  mutate(date_time = format(date_time, format = "-%Y-%m-%d %H"))

hourly_calories <- hourly_calories %>%
  separate(date_time, into = c("date", "time"), sep = " ") %>%
  mutate(date = ymd(date)) 

```

```{r,echo=FALSE, warning=FALSE}
hourly_calories %>%
  group_by(time) %>%
  summarise(average_calories = mean(calories)) %>%
  ggplot(mapping = aes(x = time, y = average_calories, fill = average_calories)) +
  geom_col() +
  labs(title = "Calories burnt throughout the day", x = "", y = "", fill = "Average Calories") +
  theme(axis.text.x = element_text(angle = 0))+
  ylim(-20,130) +
  coord_polar(start = 0)+
  custom_theme()+
  scale_fill_gradient(low = "LightGreen", high = "tan1")
```

**Insight:** Everyday, 5 pm to 7 pm is the most common time of the day
when users tend to burn more calories. Users burn more than 120 calories
per hour during this time. Other than that, the calorie burning is
seemed to be steady throughout the day.There is a dramatic drop in the
calorie burning activities since evening and keeps decreasing till
midnight. During the night time, the calorie expense is steady at around
70 calories per hour.

### **CONCLUSION and SUMMARY:**

-   Amount of distance traveled by the user does not affect their sleep
    quality i.e, covering a greater distance doesn't necessarily mean
    that the user is going to have a better sleep (on average).

-   The other thing we have noticed is that slight amount of activity
    just before going to bed can help achieve a better sleep. We can use
    this insight and add a feature which can help users set a reminder
    of doing some amount of activity so that they can reach that optimal
    steps range between 5,000 and 9,999 which could help user achieve a
    better sleep quality.

-   The 33 users with different active levels have very minimal
    difference in the total number of steps in a week.

-   Most of the users are sedentary with respect to steps in minutes and
    calories burnt.

-   Counting calorie intake and burn are one of the prominent task for
    the health conscious people. Therefore we tried to get insights from
    the data on what are the factors affecting the calorie burn and how
    effective they are. Its been observed that active individuals tend
    to burn more calories than the individuals who majorly do sedentary
    activities. Also, Running is more effective in calorie burning than
    walking.

### REFERENCES:

[1]
<https://www.kaggle.com/datasets/nurudeenabdulsalaam/fitbit-fitness-tracker-data>
